% ============================================================
% DOE Genesis Mission CFP 2026 — Pre-Proposal
% PA-AKG: Provenance-Aware Agentic Knowledge Graphs
% Focus Area: 18E — Trustworthy AI for Scientific Software
% ============================================================

\documentclass{files/proposalnsf}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{microtype}
\usepackage{enumitem}
\setlist{nosep, leftmargin=1.5em, topsep=0.2em}

\input{files/code_world}

\begin{document}

\begin{center}
    \large{\proposalEighteen} \\ \vspace{0.5pt}
\footnotesize{PI: Chris Brown, Computer Science, Virginia Tech} \\
\footnotesize{Co-PI: Xuan Wang, Computer Science, Virginia Tech} \\
\footnotesize{Co-PI: Scott T. Steinmetz, Sandia National Labs} \\
\footnotesize{GRA: Jason Cusati, Computer Science, Virginia Tech}
\end{center}

\small{\noindent\textbf{Focus Area:} 18 -- HPC Code Curation, Translation, and Development for Accelerated Scientific Discoveries} \hfill \\
\small{\textbf{Topic:} E. Trustworthy AI for Scientific Software}

\subsection*{Project Summary / Abstract}

Scientific software is the backbone of DOE's research enterprise~\cite{}, encoding decades of expertise in numerical methods, simulation algorithms, and HPC optimizations across national laboratory facilities. Recent advances in large language models (LLMs) and generative AI have introduced powerful tools for code assistance---capable of suggesting optimizations, detecting issues, and translating legacy codebases across hardware targets~\cite{}. Yet AI systems assisting with HPC scientific software cannot provide the trustworthiness guarantees these applications require. They may suggest code transformations but cannot trace each recommendation to the specific performance benchmarks, numerical analyses, or validation results that justify it~\cite{Ji2023HallucinationSurvey}. Without this traceability, AI-assisted code changes cannot be audited or reproduced---and in DOE scientific software, numerically incorrect changes can silently invalidate entire simulation campaigns.

To address these challenges, we will introduce \textbf{provenance-aware agentic knowledge graphs} (PA-AKG) as foundational infrastructure for trustworthy AI-assisted scientific software development. A persistent knowledge graph (KG) encodes HPC software knowledge---algorithms, numerical methods, performance benchmarks, hardware constraints, and validation results---linked by explicit semantic relations, providing an auditable substrate for AI code recommendations. As AI agents suggest transformations, an automated extraction layer produces \textit{span-level evidence alignment}, linking each recommendation to the specific benchmark, literature passage, or test outcome that supports it. Confirmed decisions are written back to the KG, accumulating an auditable development history. PA-AKG will offer a significant AI advantage over existing code assistance tools, enabling provenance tracking, reproducibility, and uncertainty quantification as architectural guarantees rather than post-hoc aspirations.

The proposed work will design and implement the PA-AKG infrastructure for HPC scientific software contexts, incorporating AI directly into code curation, translation, and validation workflows. We will evaluate PA-AKG against existing approaches, investigating the provenance completeness, reproducibility, and trustworthiness of AI-assisted code development. This project will advance the foundations for trustworthy AI in scientific software by designing techniques to enforce evidence traceability at every layer of the agentic architecture. This work will produce deployable tools, including an open-source PA-AKG implementation, HPC-focused benchmarks, and empirical insights to lay the groundwork for a new generation of trustworthy AI-augmented scientific software development.

\newpage
\pagenumbering{arabic}

\begin{center}
    \large{\proposalEighteen} \\ \vspace{0.5pt}
\footnotesize{PI: Chris Brown, Computer Science, Virginia Tech} \\
\footnotesize{Co-PI: Xuan Wang, Computer Science, Virginia Tech} \\
\footnotesize{Co-PI: Scott T. Steinmetz, Sandia National Labs} \\
\footnotesize{GRA: Jason Cusati, Computer Science, Virginia Tech}
\end{center}\vspace{-15pt}

\subsection*{Project Narrative}

\paragraph*{Introduction}
The DOE national laboratory system maintains some of the most complex and consequential scientific software in the world~\cite{}, encoding decades of expertise in numerical methods, simulation algorithms, and HPC architectures across facilities such as Argonne National Laboratory, Oak Ridge National Laboratory, and Sandia National Laboratories. These codebases must evolve continuously---porting to new GPU architectures, adapting to new hardware constraints, incorporating new numerical methods~\cite{}---while preserving the correctness and reproducibility that scientific integrity demands. This widening \textbf{software trustworthiness gap} poses significant challenges for national research priorities: the complexity of HPC scientific software outpaces the capacity of human reviewers to manually verify every AI-assisted change, creating critical accountability risk for DOE's scientific mission.

Recent advances in large language models (LLMs) and generative AI have produced tools of remarkable capability to assist with scientific software development. AI-based systems can suggest code optimizations, identify performance bottlenecks, and translate legacy codebases at significant scale~\cite{}. The potential of LLM-based agents to accelerate HPC software development is widely recognized and actively being pursued across the national laboratory system~\cite{rummler2026sandians}. However, LLMs are prone to hallucinate---generating plausible-sounding recommendations that conflate numerically distinct algorithms, suggest transformations that do not preserve correctness, or cite validation results that do not support the proposed optimization~\cite{Ji2023HallucinationSurvey}. These outcomes can corrupt production scientific software and invalidate research results. Recently, agentic systems coordinating multiple AI agents~\cite{} and augmentation methods to retrieve relevant documentation~\cite{} aim to ground code recommendations in existing knowledge. However, fundamental challenges that constrain AI utility for trustworthy HPC software development remain:

\textbf{Absence of persistent knowledge state.} Current AI code assistance systems treat each interaction independently, with no persistent representation of a codebase's development history, performance characteristics, or the rationale behind prior decisions. This prevents AI from reasoning about accumulated software knowledge, detecting regressions against established benchmarks, or avoiding repeated analysis of already-validated transformations.

\textbf{Lack of traceable provenance.} Trustworthy software engineering requires that decisions be justifiable and reproducible~\cite{}. When an AI system recommends a code transformation, developers must be able to trace that recommendation to specific benchmarks, numerical analyses, or validation results. Current systems may surface relevant documentation but do not provide structured reasoning that connects evidence to recommendations---making AI-assisted code changes impossible to audit or reproduce independently. \vspace{-5pt}


\problem{The lack of structured and persistent knowledge about HPC software development limits AI agents' ability to provide trustworthy, reproducible assistance for DOE scientific software.}{}

To address these challenges, this project proposes the design, implementation, and evaluation of \textbf{provenance-aware agentic knowledge graphs} (PA-AKG) for HPC scientific software development. PA-AKG is foundational infrastructure for trustworthy, AI-integrated software engineering, addressing the challenges above by providing AI agents with a persistent, structured, and auditable representation of the current state of a scientific software system in a given domain. Knowledge graphs (KGs) have been applied in software engineering and scientific computing as a mechanism for representing complex, heterogeneous information through explicit semantic relationships~\cite{}. PA-AKG will encode HPC software knowledge (\ie algorithms, numerical methods, performance benchmarks, hardware constraints, validation results, and development decisions) as nodes representing existing knowledge, with relationships between them as directed edges. As AI agents suggest code transformations, they compare proposed changes against KG-encoded performance expectations. When decisions are confirmed, they are written back to the KG with explicit provenance chains---carrying the specific benchmarks and evidence that justify each change---accumulating an auditable development history that future agents can build upon.

\paragraph{Objectives.} Our goal is to design and implement a provenance-first agentic KG architecture with three integrated layers: knowledge representation; automation and extraction; and agentic orchestration. We will implement the PA-AKG architecture, developing infrastructure to automate the extraction and structuring of claim-evidence pairs in HPC software contexts with role-specific agents (\eg ranking agents, evaluation agents, and synthesis agents). The initial implementation will focus on the software engineering research domain, following our prior work~\cite{cusati2026fose}, with secondary validation against HPC performance analysis workflows at Sandia National Laboratories. Software engineering, the processes and methods for developing and maintaining software systems, is also critical for scientific discovery and innovation~\cite{vos2026reclaiming}, enabling infrastructure across domains including next-generation computing, national security, and biotechnology. We will conduct a comprehensive evaluation of our approach, comparing PA-AKG against RAG, GraphRAG~\cite{Edge2024GraphRAG}, and static baseline approaches on provenance completeness, reproducibility, and trustworthiness of AI-generated software decisions. We will also conduct controlled user studies with HPC software developers to assess usability and utility. Key risks include extraction reliability at HPC code scale and KG schema portability across software domains; these are mitigated by the staged evaluation design and Sandia's established HPC validation corpora.

\paragraph{Research Plan.} Our research plan involves four main phases:

\textit{Phase 1 (Months 1--3): Knowledge Representation Layer.}
We will design the HPC software KG schema with entity types such as \textsc{Algorithm}, \textsc{NumericalMethod}, \textsc{CodeComponent}, \textsc{HardwareConstraint}, \textsc{Benchmark}, and \textsc{ValidationResult}---linked by semantic relations (\textsc{implements}, \textsc{requires}, \textsc{supersedes}, \textsc{validates}). We will conduct pilot studies (\eg interviews with HPC software developers) to obtain insights from practitioners, informing entity types and relationships for our initial implementation. A hybrid symbolic--semantic design pairs a property graph store with a vector index, supporting both structured relational queries and approximate similarity search. We will populate an initial KG from the software engineering research corpus established in prior work, validate entity extraction against manually annotated ground-truth annotations, and establish baseline KG coverage metrics.

\textit{Phase 2 (Months 4--6): Automation, Extraction, and Agentic Orchestration.}
We will implement the extraction and automation pipeline: structured prompting will produce claim--evidence pairs; span-level alignment will link each claim to character-level source passage spans; and canonicalization auditing will normalize, validate, and deduplicate entity references to canonical KG nodes. The agentic orchestration layer, adapting the Planning \& Control pattern~\cite{Park2023GenerativeAgents}, introduces four role-specific agent types. \textit{Ranking agents} score candidate code changes by evidence strength and risk. \textit{Continuation agents} propose follow-on validation analyses grounded in KG gaps. \textit{Evaluation agents} execute reproducible benchmark workflows and write results back to the KG as structured evidence nodes. \textit{Synthesis agents} generate recommendation artifacts---proposed transformation, supporting evidence chain, predicted performance impact---with full provenance metadata. The provenance inheritance constraint is enforced at every agent handoff, ensuring no reasoning step discards the justification for prior decisions.

\textit{Phase 3 (Months 7--9): Evaluation and Generalization.}
We will conduct two comprehensive evaluations of our PA-AKG approach. The first will benchmark our system against existing approaches, including vector-based RAG, hybrid retrieval, and GraphRAG-style reasoning. We will construct a benchmark of human-annotated claim--evidence pairs from the SE research corpus. We will calculate span-level alignment precision/recall, Cohen's $\kappa$ with expert-labeled canonical entity clusters, and traceability completeness (\ie \% of KG nodes with complete provenance chains). A secondary HPC validation at Sandia National Laboratories will assess cross-domain generalizability of the extraction pipeline. Our second evaluation will assess provenance completeness and trustworthiness of AI-assisted code recommendations through controlled user studies with HPC software practitioners, collecting quantitative task-completion metrics and qualitative feedback on recommendation auditability.

\textit{Phase 4: Dissemination.}
We will pursue targeted dissemination of PA-AKG research findings to academic and practitioner audiences. Academic outputs will be submitted to premier venues in software engineering, AI, and machine learning, including the International Conference on Software Engineering (ICSE)~\cite{}, the Mining Software Repositories conference (MSR)~\cite{}, and the Supercomputing Conference (SC)~\cite{}. We will submit to tool, demo, and tutorial tracks at these venues to provide hands-on exposure of PA-AKG to HPC software development and software engineering research communities, lowering the barrier to adoption across DOE and academic research programs. To reach the practitioner community, we will release the PA-AKG implementation as an open-source package with comprehensive documentation, worked examples, and integration guides targeting national laboratory developers and the broader scientific computing community. We will present findings at Sandia National Laboratories and partner institutions to gather practitioner feedback, surface deployment barriers, and assess adoption pathways within active DOE HPC workflows. This phase will conclude with an impact assessment evaluating how PA-AKG changes developers' ability to audit and reproduce AI-assisted code decisions---informing the design of future trustworthy AI systems for HPC scientific software and establishing PA-AKG as a reference model for accountable AI assistance in DOE's software engineering mission.

\paragraph{Expected Outcomes.} The expected outcomes include:
\begin{itemize}
    \item A design for provenance-aware knowledge graph architecture for trustworthy AI-assisted HPC scientific software development.
    \item An open-source PA-AKG implementation, consisting of an automated extraction pipeline and agentic orchestration, applied to HPC software development contexts.
    \item Benchmarking infrastructure to assess LLM-based systems for trustworthy code assistance in scientific software.
    \item Empirical insights on the provenance completeness, reproducibility, and trustworthiness of AI-assisted HPC code decisions.
    \item Broader scientific impact through open engagement with DOE national laboratory and academic research communities, informing the design of next-generation trustworthy AI tools for HPC scientific software and establishing PA-AKG as a reference model for accountable AI in DOE's software mission.
\end{itemize}

\paragraph{Research Team.} This project aligns with the prior work and trajectory of the research team.
\textbf{PI Brown} and GRA Cusati explored how LLMs propagate research-based evidence~\cite{brown2025esem} and outlined researchers' desire for structured knowledge accumulation~\cite{cusati2026fose} in software engineering contexts. For this work, Dr. Brown will lead the overall execution of the project---including the PA-AKG implementation and research plan. \textbf{Co-PI Wang} investigates LLMs in the context of science and society, with substantial experience studying the applications of LLMs (\ie~\cite{bi2024ai,srivastava2025llmthinkbench}) and agents (\ie~\cite{lu2024triageagent,latimer2025hindsight}) in scientific settings and earning an NSF CAREER award to extract patterns from research data~\cite{wang_career}. Dr. Wang will inform the design of the PA-AKG architecture, particularly focusing on knowledge retrieval and extraction, multi-agent orchestration, and knowledge persistence, and the benchmark evaluation~\cite{srivastava2025beyondbench}. \textbf{Co-PI Steinmetz} is a researcher at Sandia National Labs with expertise in human-machine teaming~\cite{chance2023towards,powell2021coordination}, generative AI~\cite{krofcheck2023overview,layton2025mstd}, and trust in AI systems~\cite{steinmetz2025trust,naugle2025trusted}. Dr. Steinmetz will inform the PA-AKG design and support validation efforts against HPC software development workflows at Sandia, leveraging his expertise in trustworthy AI systems to assess usability and accountability in practical HPC development contexts. \textbf{GRA Cusati} is a Ph.D. student in the Department of Computer Science at Virginia Tech with extensive industry experience implementing complex production software in various roles. Jason will lead the design and development of the PA-AKG infrastructure and the overall data collection and analysis efforts.

\noindent\textbf{Resources.} Virginia Tech's Advanced Research Computing~\cite{} provides GPU cluster infrastructure for large-scale KG construction and LLM inference workloads. The team will access publicly available scientific software corpora and documentation (\eg GitHub, arXiv, IEEE Xplore) and pre-trained language models (\eg Llama~\cite{}, GPT-4~\cite{}) for extraction and orchestration. Sandia National Laboratories will contribute access to HPC performance benchmarks, legacy scientific codebases, and DOE-relevant software documentation corpora for secondary domain validation, ensuring the research addresses realistic national laboratory software development challenges.

\bibliographystyle{IEEEtran}
\bibliography{references,../references}

\end{document}
