% ============================================================
% Data Sources and Models — proposal_18E/data_sources.tex
% Reviewer concern: RF-5b (Data Sources TODO)
% Verified by: AC-6
%
% This file currently reflects PLAN B (public corpora) as a safe default.
% PLAN A (specific Sandia Kokkos/Trilinos/LAMMPS corpus names) will be
% swapped in after the 2026-04-09 Sandia meeting — see \todo{} marker below.
% ============================================================

% [RF-5b] Data Sources — Plan B default, Plan A specifics pending Sandia meeting — verified by: AC-6

\paragraph{Knowledge graph corpus.} We will devise a corpus from prior work extended with HPC-focused publications from arXiv (cs.DC, cs.SE)~\cite{arxiv}, IEEE Xplore~\cite{ieee}, and the ACM digital library~\cite{acm}---seeding the initial calibration (Phase 1) and expanded corpus (Phase 2) described in \S\ref{sec:methods}.

\paragraph{HPC codebases and performance data.} We will anchor the HPC evaluation on Sandia-based scientific software systems that together exercise the full performance-portability-to-application stack. The initial implementation focuses on \textbf{Pyomo}~\cite{hart2017pyomo}---a Python-based open-source package originally developed at Sandia National Laboratories. Pyomo supports formulating and solving a broad class of optimization problems, including linear, quadratic, nonlinear, mixed-integer (linear, quadratic, nonlinear, and stochastic), generalized disjunctive, differential algebraic equation, and mathematical programming with equilibrium constraint formulations. The breadth of problem types, layered abstraction model, integration in research initiatives, and extensive history of algorithmic design decisions accumulated through active Sandia development make Pyomo a well-instrumented corpus for PA-AKG---encoding complex, heterogeneous scientific knowledge (\eg numerical method choices, solver compatibility trade-offs, performance characteristics, etc.) that can be explicitly represented in a KG for trustworthy AI-assisted code recommendations. The broader evaluation (Phase 3) will target additional codebases for cross-domain generalizability, including \textbf{Kokkos} (performance portability layer), \textbf{Trilinos} (numerical solver framework), and \textbf{LAMMPS} (molecular dynamics application).

% Public data sources include the Extreme-Scale Scientific Software Stack (E4S) ecosystem, the public LAMMPS GitHub history and pair-style benchmark corpus, public Kokkos tutorials and regression tests, the Trilinos open benchmark suite, and the Pyomo GitHub repository history and associated benchmark artifacts. Performance and compiler-trace data will be drawn from SPEC CPU/HPC benchmarks as a public baseline.
% \todo{RF-5b (post-Sandia-meeting): after the 2026-04-09 meeting with Co-PI Steinmetz, swap in the specific Sandia Pyomo/Kokkos/Trilinos/LAMMPS corpus subsets, benchmark logs, and validation artifacts accessible under existing Sandia data agreements. Cite the data agreement and name the specific code subsets.}

\paragraph{Pre-trained models.} For local inference and privacy-sensitive prototyping we will use \textbf{Llama 3} (Meta, open-weight). For comparison and high-capability benchmarks we will evaluate \textbf{GPT-4} (OpenAI API) and \textbf{Claude} (Anthropic API). For code-specific tasks, \textbf{CodeLlama} provides a code-pretrained baseline.

\paragraph{Graph storage and retrieval infrastructure.} The PA-AKG substrate will be implemented on \textbf{Neo4j} (primary development) with Amazon Neptune evaluated as a cloud-deployment target. The dense-retrieval vector index will use \textbf{FAISS} for local experimentation and \textbf{pgvector} for the integrated property-graph + vector-index queries required by the canonicalization-auditing pipeline.

\paragraph{Data access.} Virginia Tech has existing access to public corpora (arXiv, IEEE Xplore, GitHub). Sandia National Laboratories, via Co-PI Steinmetz, provides access to the HPC validation corpora identified above under existing Sandia data agreements. The specific data-transfer mechanism and scope will be finalized in coordination with Sandia information-release procedures.

\paragraph{Resources.} The Department of Computer Science at Virginia Tech is equipped to support the proposed research. Virginia Tech's Advanced Research Computing~\cite{arc} and Computer Science Research Virtual Machine cluster (CSRVM) will provide GPU cluster infrastructure and computing resources for large-scale KG construction and LLM inference workloads. Sandia National Laboratories will contribute access to HPC performance benchmarks, legacy scientific codebases, and DOE-relevant software documentation for secondary domain validation. More details are provided in Appendix~\ref{app:feors}.

% , ensuring the research addresses realistic national laboratory software development challenges