% ============================================================
% Decision Gate Metrics — proposal_18E/decision_gate.tex
% Reviewer concern: RF-5a (Decision Gate Metrics TODO)
% Verified by: AC-5
%
% Framed as SCALING-BEHAVIOR CLAIMS (not fixed-number commitments).
% Thresholds are discovered empirically via the calibration studies
% in Methods §Phase 1 (AC-5b) and §Phase 2 (AC-5c).
% ============================================================

% [RF-5a] Decision Gate Metrics as scaling-behavior claims with empirically-discovered thresholds — verified by: AC-5


% \paragraph{Empirically-discovered thresholds.} Rather than pre-committing to first-pass numeric targets, the Phase I go/no-go thresholds for provenance completeness, reproducibility rate, auditability index, and span-level alignment F1 are \emph{discovered} through the calibration studies in \S\ref{sec:methods}. 

% \paragraph{Iterate-if-low rule.} If the Phase 1 or Phase 2 calibration study produces thresholds below a defensible trustworthiness floor---operationally, below the point at which human HPC reviewers consistently judge recommendations as auditable---the team will iterate the PA-AKG implementation (schema, extraction pipeline, or provenance inheritance rules) and re-calibrate before the Month-6 gate, rather than accepting a weak result as the project's claimed AI advantage. This rule protects the integrity of the scaling-behavior claim: improvements must come from additional curated data, not from lowering the bar.

\paragraph{Trustworthiness measurement.} We will assess the reliability of AI-generated code recommendations by quantifying their success rate in passing validation tests against manual annotation and established benchmarks. This metric aims for at least 90\% of AI-assisted modifications to meet predefined correctness criteria. By ensuring that a high percentage of AI-generated suggestions align with expected outcomes, this measurement instills confidence in both developers and stakeholders regarding the efficacy of AI integration in scientific software. Trust in these recommendations is essential for wider adoption and further investment in AI-assisted development processes within the DOE's scientific missions.

\paragraph{Auditability measurement.} We will also measure the depth and clarity of provenance tracking associated with each code recommendation proposed by AI agents. This metric quantifies how easily each modification can be traced back to specific performance benchmarks, literature references, and validation results, facilitating an auditable development history. An auditability Index score of 95\% or higher would indicate that a significant majority of code changes are thoroughly documented, ensuring that developers can verify and trust the rationale behind AI recommendations. This provenance is vital for maintaining scientific integrity and enabling independent validation of results, thereby addressing critical trustworthiness issues.

\paragraph{Provenance-driven acceleration.} We will investigate whether PA-AKG's persistent knowledge layer creates direct acceleration, as AI agents resolve queries from KG-stored validated knowledge without re-executing the full retrieval and inference pipeline. We will target $\geq$30\% reduction in mean response latency for transformation queries with encoded answers validated provenance in the KG, compared to baseline RAG systems that treat every query independently. This metric will capture the compounding benefit of PA-AKG's knowledge state. As the KG grows, the fraction of queries answered from validated provenance increases, reducing the lag between new research findings and actionable code optimizations.

\paragraph{AI advantage as scaling behavior.} We will assess the monotonic improvement of provenance completeness, trustworthiness, efficiency, and span-level alignment F1 as the underlying PA-AKG corpus scales from the Phase 1 small-seed calibration ($\sim$5--10 papers) to the Phase 2 expanded-corpus calibration ($\sim$50+ papers). Phase 1 calibration produces an initial threshold set published at the end of Month 3; Phase 2 re-calibration re-validates and refines these against the expanded corpus in Month 5. This threshold comparison will provide direct evidence that our approach translates into measurable improvements in the trustworthiness of PA-AKG's outputs at scale, directly addressing the call for ``scaling behavior which shows increasing performance as additional data, computing, and/or other resources are applied''.


\paragraph{Supporting user study signals.} As an early signal, controlled studies with HPC software developers and researchers will assess recommendation auditability, with a target of $\geq$70\% of developers rating PA-AKG recommendations as auditable, compared against baseline LLM recommendations. We will also track the amount of time for developer verification and decision making for AI-generated recommendations, aiming to achieve a measurable reduction in time spent on study tasks with improved accuracy compared to baseline LLM-supported workflows. Additional qualitative feedback from user study participants on usability and integrability within R\&D workflows will provide supporting insights on a potential Phase II decision.