% ============================================================
% DOE Genesis Mission CFP 2026 — Pre-Proposal
% PA-AKG: Provenance-Aware Agentic Knowledge Graphs
% Focus Area: 11C — AI-Accelerated Science: Correlation to Understanding
% ============================================================

\documentclass{files/proposalnsf}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{microtype}
\usepackage{cite}
\usepackage{enumitem}
\setlist{nosep, leftmargin=1.5em, topsep=0.2em}

\input{files/code_world}

\begin{document}

\begin{center}
    \large{\proposalEleven} \\ \vspace{0.5pt}
\footnotesize{PI: Chris Brown, Computer Science, Virginia Tech} \\
\footnotesize{Co-PI: Xuan Wang, Computer Science, Virginia Tech} \\
\footnotesize{Co-PI: Scott T. Steinmetz, Sandia National Labs} \\
\footnotesize{GRA: Jason Cusati, Computer Science, Virginia Tech}
\end{center}

\small{\noindent\textbf{Focus Area:} 11 -- Achieving AI-Driven Autonomous Laboratories} \hfill \\
\small{\textbf{Topic:} C. AI-Accelerated Science: Correlation to Understanding}

\subsubsection*{Project Summary / Abstract}

The Genesis Mission calls for AI that can close the loop between data and discovery: identifying phenomena in high-throughput scientific data that are inconsistent with current models and generating plausible, testable hypotheses as explanations~\cite{EO14363}. This is the ``correlation to understanding'' challenge. Current AI systems can retrieve relevant documents and generate plausible text, but they cannot systematically compare new observations against an explicit, persistent representation of current scientific knowledge, nor produce hypotheses with traceable provenance back to the specific evidence that supports each inference~\cite{Ji2023HallucinationSurvey}. Without persistent knowledge infrastructure, every agent session starts from scratch---unable to build on prior findings, detect genuine anomalies against an accumulated model, or produce outputs that human researchers can audit and extend.

The vision of this project is \textbf{provenance-aware agentic knowledge graphs as foundational infrastructure for closed-loop AI-accelerated scientific discovery}. A persistent research knowledge graph (KG) encodes the current state of scientific understanding as first-class entities---problems, datasets, methods, constraints, and evidence---linked by explicit semantic relations~\cite{Hogan2021KnowledgeGraphs}. This structured representation is the ``current model'' against which incoming scientific data is continuously compared. As new evidence arrives from literature, experiments, or simulations, an automated extraction layer produces \textit{span-level evidence alignment}---linking each extracted claim to the specific source passage that licenses it---and \textit{canonicalization auditing}---resolving entity references to canonical KG nodes. Agents compare new findings against KG-encoded expectations; when divergences are detected, hypothesis-generation agents produce structured, evidence-grounded explanations with explicit provenance chains. Confirmed hypotheses are written back to the KG, updating the current model and enabling the next discovery cycle. This closed loop---observe, compare, hypothesize, validate, accumulate---directly implements the 11C vision of AI integrated into scientific workflows, combining real-time analysis, intelligent feedback, and hypothesis generation.

AI enhances the scientific discovery workflow in three concrete ways that are not possible with current retrieval approaches~\cite{Lewis2020RAG}. First, \textbf{persistence}: scientific knowledge accumulates in the KG across research sessions, enabling AI to reason over an evolving model of what is known and identify genuine gaps rather than re-deriving the same correlations repeatedly. Second, \textbf{grounded anomaly detection}: by maintaining an explicit representation of current model expectations, agents can detect observations that are statistically plausible but inconsistent with established knowledge---the ``correlation'' that 11C calls for transforming into ``understanding.'' Third, \textbf{traceable hypothesis generation}: every hypothesis is accompanied by a structured provenance chain linking each inferential step back to the supporting evidence, enabling human researchers to evaluate AI-generated hypotheses against their own domain knowledge and extend rather than merely accept or reject them.

The proposed system builds directly on prior work establishing the need and feasibility of this approach~\cite{cusati2026fose, brown2025esem}. The research will be validated in the software engineering research domain, leveraging empirical standards and methodological frameworks already established in that community, with secondary validation in one additional scientific domain to assess generalizability. Expected outcomes include a reference architecture for provenance-enforcing agentic research systems, an evaluation framework with four quantitative axes (E1--E4), and a reproducibility protocol specifying the artifacts required for independent verification of AI-generated scientific findings.

\newpage
\pagenumbering{arabic}

\begin{center}
    \large{\proposalEleven} \\ \vspace{0.5pt}
\footnotesize{PI: Chris Brown, Computer Science, Virginia Tech} \\
\footnotesize{Co-PI: Xuan Wang, Computer Science, Virginia Tech} \\
\footnotesize{Co-PI: Scott T. Steinmetz, Sandia National Labs} \\
\footnotesize{GRA: Jason Cusati, Computer Science, Virginia Tech}
\end{center}

\subsubsection*{Research Plan}

\paragraph{Introduction.}
DOE's 11C challenge names a precise capability gap: AI systems today automate data collection and identify statistical correlations, but cannot autonomously determine whether a new correlation represents a genuinely novel phenomenon inconsistent with current models, and cannot generate a plausible explanation as a testable hypothesis. This gap exists because the capability requires two things that current AI systems lack together: (1) a persistent, queryable representation of the current scientific model---not just retrieved documents, but structured knowledge encoding what the field currently believes and why---and (2) a provenance mechanism that traces every inference back to the specific evidence supporting it, enabling both automated comparison and human verification.

Semi-autonomous agentic systems now coordinate multiple AI agents, invoke external tools, and execute multi-step research workflows with limited human oversight~\cite{Park2023GenerativeAgents}. Yet their outputs lack the infrastructure for persistence, traceability, and cumulative knowledge progression. Without \textit{persistence}, research knowledge accumulated in one session decays before the next. Without \textit{traceability}, generated claims cannot be verified against their source evidence. Without \textit{evaluability}, there is no systematic way to assess whether outputs are improving or degrading over time~\cite{cusati2026fose}.

These three infrastructure requirements form a cycle: without persistent knowledge infrastructure, provenance tracking is infeasible; without provenance, rigorous evaluation is impossible; without evaluation criteria, architectural design lacks a target to optimize against. The proposed research breaks this cycle by addressing provenance and architectural integration simultaneously. Our prior work establishes both the problem and the approach: Cusati \& Brown~\cite{cusati2026fose} identify the three structural gaps in current AI research assistance systems and propose the provenance-first agentic KG as the solution; Brown \& Cusati~\cite{brown2025esem} demonstrate empirically that current AI tools propagate evidence inconsistently, establishing the practical urgency of span-level alignment.

\paragraph{Objectives.}
\begin{enumerate}
    \item Design and implement a provenance-first agentic KG architecture with three integrated layers: knowledge representation, automation and extraction, and agentic orchestration.
    \item Implement span-level evidence alignment---automated extraction producing structured claim--evidence pairs with source passage linkage---as an architectural invariant enforced across all agent operations.
    \item Implement canonicalization auditing---entity normalization, validation, and deduplication to canonical KG nodes---enabling consistent cross-session entity resolution.
    \item Build role-specific agent orchestration: ranking agents, continuation agents, evaluation agents, and synthesis agents operating under provenance inheritance constraints.
    \item Evaluate the system along four quantitative axes (E1--E4) against RAG, GraphRAG~\cite{Edge2024GraphRAG}, and static snapshot baselines, measuring improvements in persistence, traceability, and hypothesis quality.
\end{enumerate}

\paragraph{Research Questions.}
\begin{itemize}
    \item \textbf{RQ1 (E1):} Does span-level evidence alignment improve claim-to-source traceability over unstructured RAG baselines, measured by precision and recall against human-annotated gold spans?
    \item \textbf{RQ2 (E2):} Does canonicalization auditing produce entity resolutions consistent with human expert judgment, measured by Cohen's $\kappa$ against expert-labeled canonical clusters?
    \item \textbf{RQ3 (E3):} Does persistent KG infrastructure support cumulative research progression across sessions, measured by traceability completeness (fraction of nodes with complete provenance chains)?
    \item \textbf{RQ4 (E4):} Does the KG's provenance completeness and hypothesis quality remain stable or improve as the research corpus evolves over time?
\end{itemize}

\paragraph{Detailed Research Plan.}

\textit{Phase 1 (Months 1--3): Knowledge Representation Layer.}
We will design the research KG schema with first-class entity types---\textsc{Problem}, \textsc{Dataset}, \textsc{Method}, \textsc{Constraint}, \textsc{Evidence}---linked by semantic relations (\textsc{supports}, \textsc{contradicts}, \textsc{enables}, \textsc{requires}). Research problems are modeled as tractable, scoped questions with identifiable assumptions, methods, and evaluation criteria, extracted from source documents with assumptions, constraints, datasets, metrics, and evidence spans anchored to source passages. A hybrid symbolic--semantic design pairs a property graph store with a vector index, supporting both structured relational queries and approximate similarity search. We will populate an initial KG from the software engineering research corpus established in prior work~\cite{cusati2026fose, brown2025esem}, validate entity extraction against manually annotated ground-truth annotations, and establish baseline KG coverage metrics.

\textit{Phase 2 (Months 4--6): Automation, Extraction, and Agentic Orchestration.}
We will implement the extraction pipeline: structured prompting produces claim--evidence pairs; span-level alignment links each claim to character-level source passage spans; canonicalization auditing normalizes, validates, and deduplicates entity references to canonical KG nodes. The agentic orchestration layer, adapting the Planning \& Control pattern~\cite{Park2023GenerativeAgents}, introduces four role-specific agent types. \textit{Ranking agents} score candidate research directions by tractability and evidence strength. \textit{Continuation agents} propose follow-on analyses grounded in KG gaps. \textit{Evaluation agents} execute reproducible experimental workflows and write results back to the KG as structured evidence nodes. \textit{Synthesis agents} generate hypothesis artifacts---problem statement, supporting evidence chain, proposed test, predicted outcome---and curate them to the KG as \textsc{Hypothesis} nodes with full provenance metadata. The provenance inheritance constraint is enforced at every agent handoff: each agent receives a provenance context object and may only produce outputs that extend, not discard, the inherited evidence chain.

\textit{Phase 3 (Months 7--9): Evaluation and Generalization.}
We will evaluate along the E1--E4 framework. \textbf{E1}: span-level alignment precision/recall vs. 500 human-annotated claim--evidence pairs from the SE corpus. \textbf{E2}: canonicalization Cohen's $\kappa$ vs. expert-labeled canonical entity clusters. \textbf{E3}: traceability completeness (\% KG nodes with complete provenance chains) as a function of corpus size. \textbf{E4}: longitudinal stability score across six simulated corpus update cycles. Baselines: (B1) vector RAG without span alignment; (B2) GraphRAG~\cite{Edge2024GraphRAG} without provenance enforcement; (B3) static snapshot system without cumulative KG. A secondary scientific domain will be evaluated to assess cross-domain generalizability of the extraction pipeline and schema.

\paragraph{Team Roles.}
\textbf{PI Chris Brown} leads project direction, KG architecture design, and evaluation methodology (E1--E4). \textbf{Co-PI Xuan Wang} leads multi-agent orchestration design, ranking and continuation agent implementation, and embedding-based canonicalization. \textbf{Co-PI Scott T. Steinmetz} (Sandia National Labs) leads provenance infrastructure design and secondary domain validation against DOE scientific corpora. \textbf{GRA Jason Cusati} implements the system end-to-end, constructs and annotates the evaluation corpus, and executes experimental evaluation and baseline comparisons.

\bibliographystyle{IEEEtran}
\bibliography{references}

\end{document}